---
output:
  pdf_document: default
  html_document: default
---
# Reliability and Validity

For this chapter we refer to the book 
[Measurement in Medicine](https://www.cambridge.org/core/books/measurement-in-medicine/8BD913A1DA0ECCBA951AC4C1F719BCC5).

I invite you to read the introductory chapters 1 and 2 about concepts,
theories and models, and types of measurement.

In general, when conducting a measurement of any sort 
(laboratory measurements, scores from questionnaires, etc.), 
we want to be reasonably sure

- that we actually **measure what we intend to measure**; 
  ([validity](https://en.wikipedia.org/wiki/Validity_(statistics)); 
  chapter 6 in the book);
- that the measurement does **not change too much** if the 
  underlying **conditions are the same**
  ([reliability](https://en.wikipedia.org/wiki/Reliability_(statistics)); 
  chapter 5 in the book); and
- that we are able to detect a **change** if the underlying conditions change 
  ([responsiveness](https://tinyurl.com/3vdcxy49); chapter 7 in the book); and
- that we understand the meaning of a change in the measurement
  (interpretability; chapter 8 in the book).

In this [video](https://www.youtube.com/watch?v=KuT2n1w0Ixc&ab_channel=Physiotutors), 
Kai jump starts you on reliability and validity.


## Reliability
You can watch this [video](https://www.youtube.com/watch?v=9HSoWaRpcys&ab_channel=Physiotutors) 
to get started.

Imagine, you measure a patient (pick your favorite measurement), for example,
the range of motion (ROM) of the shoulder. 

- If you are interested in how
  similar your measurements are in comparison to your colleagues, you are 
  trying to determine the so-called **inter-rater reliability**.
- If you are interested in how similar your measurements are when you measure
  the same patient twice, you are trying to determine the so-called 
  **intra-rater reliability**.

Assuming there is a true (but unknown) underlying value (of Range of Motion, ROM), 
it is clear that measurements will not be *exactly* the same. 
Possible influences (potentially) causing different results are:

- the measurement instrument itself,
- the patient (e.g., mood/motivation),
- the examiner (e.g., mood/focus, influence on patient, interpretation of decision rules),
- the environment (e.g., the room temperature).

Note that the **true score** is defined in our context as the **average**
of all measurements if we would measure repeatedly an infinite number of times (p.98).
In this sense, one can measure the **true score** (in principle, not in practice) 
arbitrarily well.

### Peter and Mary's ROM measurements

The data can be found [here](http://www.clinimetrics.nl/answers-to-the-assignments-in-textbook_22_0.html).
We randomly select 50 measurements from Peter and Mary in 50 different patients,
plot their measurements and annotate the absolutely largest one(s). At first the
*not affected shoulder* (nas) and then the *affected shoulder* (as).

```{r}
library(pacman)
p_load(tidyverse, readxl)

# Read file
url <- "https://raw.githubusercontent.com/jdegenfellner/Script_QM2_ZHAW/main/data/chapter%205_assignment%201_2_wide.xls"
temp_file <- tempfile(fileext = ".xls")
download.file(url, temp_file, mode = "wb")  # mode="wb" is important for binary files
df <- read_excel(temp_file)

head(df)
dim(df)

# As in the book, let's randomly select 50 patients.
set.seed(123)
df <- df %>% sample_n(50)
dim(df)

# "as" = affected shoulder
# "nas" = not affected shoulder

df <- df %>%
  dplyr::mutate(diff = abs(ROMnas.Peter - ROMnas.Mary))  # Compute absolute difference

max_diff_point <- df %>%
  dplyr::filter(diff == max(diff, na.rm = TRUE))  # Find the row with the max difference

df %>%
ggplot(aes(x = ROMnas.Peter, y = ROMnas.Mary)) +
  geom_point() + 
  geom_point(data = max_diff_point, aes(x = ROMnas.Peter, y = ROMnas.Mary), 
             color = "blue", size = 4) +  # Highlight max difference point
  geom_abline(intercept = 0, slope = 1, color = "red") +
  theme_minimal() +
  ggtitle("ROMnas.Peter vs. ROMnas.Mary") +
  theme(plot.title = element_text(hjust = 0.5)) +
  annotate("text", x = max_diff_point$ROMnas.Peter, 
           y = max_diff_point$ROMnas.Mary, 
           label = paste0("Max Diff: ", round(max_diff_point$diff, 2)), 
           vjust = -1, color = "blue", size = 4)

# average abs. difference:
mean(df$diff, na.rm = TRUE) # 7.2

cor(df$ROMnas.Peter, df$ROMnas.Mary, use = "complete.obs") 
```

The red line represents the line of equality ($y=x$). If the measurements are 
exactly the same,
all points would lie on this line ($\rho=1$). The blue point represents the measurement with the largest 
absolute difference in measured Range of Motion (ROM) values between Peter and Mary
from the randomly chosen 50 people. Note that the maximum difference in 
all 155 patients is even higher with 35 degrees.

The first simple measure of agreement we could use is the [(Pearson) correlation](https://en.wikipedia.org/wiki/Pearson_correlation_coefficient), 
which measures the **strength and direction of a linear relationship between two variables**.
But correlation does not exactly measure what we want. If there was a bias 
(e.g., Mary systematically measures 5 degrees more than Peter), correlation would not 
notice this ([exercise 1](#exercise1)). It actually is too optimistic about 
the agreement since it only cares about the linearity and not about a potential bias:
$r=0.2403213$ which indicates a weak positive correlation. Higher values of Peter's measurements
are associated with higher values of Mary's measurements.
But: Knowing Peter's measurement does not help us to 
[*predict*](https://en.wikipedia.org/wiki/Pearson_correlation_coefficient#Interpretation_of_the_size_of_a_correlation) 
Mary's measurement at such a low correlation ([exercise 2](#exercise2)).
So, on the *not affected shoulder* (nas), agreement is really bad.

**What about the affected shoulder (as)?**

```{r}
library(ggExtra)
df <- df %>%
  mutate(diff = abs(ROMas.Peter - ROMas.Mary))  # Compute absolute difference

max_diff_point <- df %>%
  dplyr::filter(diff == max(diff, na.rm = TRUE))  # Find the row with the max difference

p <- df %>%
ggplot(aes(x = ROMas.Peter, y = ROMas.Mary)) +
  geom_point() + 
  geom_point(data = max_diff_point, aes(x = ROMas.Peter, y = ROMas.Mary), 
             color = "blue", size = 4) +  # Highlight max difference point
  geom_abline(intercept = 0, slope = 1, color = "red") +
  theme_minimal() +
  ggtitle("ROMas.Peter vs. ROMas.Mary") +
  theme(plot.title = element_text(hjust = 0.5)) +
  annotate("text", x = max_diff_point$ROMas.Peter, 
           y = max_diff_point$ROMas.Mary, 
           label = paste0("Max Diff: ", round(max_diff_point$diff, 2)), 
           vjust = -1, color = "blue", size = 4)
# Add marginal histograms
ggMarginal(p, type = "density", fill = "gray", color = "black")

# average abs. difference:
mean(df$diff, na.rm = TRUE) # 7.2
cor(df$ROMas.Peter, df$ROMas.Mary, use = "complete.obs") 

# mean difference
mean(df$ROMas.Peter - df$ROMas.Mary, na.rm = TRUE) 
```

In the affected side, the average absolute difference is even larger ($7.78$)
with a maximum absolute difference of $37$ degrees (on the whole sample of $155$ patients),
but the correlation is much higher ($r=0.8516653$). See Figure 5.2 in the book.

This is an occasion for using the correlation coefficient even though
the marginal distributions are not normal: There are much more measurements in the higher
values around $80$ than below, say, $60$. But the correlation coefficient makes sense
for descriptive purposes.

In this case, knowing Peter's measurement *does* (at least somewhat) help us to predict 
Mary's measurement (see [below](#badnews)).

### Intraclass Correlation Coefficient (ICC) {#ICC}

One way to measure reliability is to use the 
[intraclass correlation coefficient (ICC)](https://en.wikipedia.org/wiki/Intraclass_correlation).

This measure is based on the idea that the observed score $Y_i$ consists of the true score 
(ROM) ($\eta_i$) and a measurement error (for each person).
The proportion of the true score variability to the total variability is the ICC.

In the background one thinks of a statistical model from the 
[Classical Test Theory (CTT)](https://en.wikipedia.org/wiki/Classical_test_theory).
There is 

- a true underlying score $\eta_i$ (for each patient $i$) and 
- an error term $\varepsilon_i \sim N(0, \sigma_i)$ which is the difference between the true
score and 
- the observed score $Y_i$.

$$Y_i = \eta_i + \varepsilon_i$$

It is assumed that $\eta_i$ and $\varepsilon_i$ are independent: ($\mathbb{C}ov(\eta_i, \varepsilon_i)=0$).
This is a convenient assumption because now we know (see [here](https://en.wikipedia.org/wiki/Variance#Addition_and_multiplication_by_a_constant)) 
that the variance
of the observed score $Y_i$ is just the sum of the variance of the true score $\eta_i$
and the variance of the error term $\varepsilon_i$:

$$\mathbb{V}ar(Y_i) = \mathbb{V}ar(\eta_i) + \mathbb{V}ar(\varepsilon_i)$$
$$\sigma_{Y_i}^2 = \sigma_{\eta_i}^2 + \sigma_{\varepsilon_i}^2$$

We want most of the variability (i.e. how much scores in our relevant population vary) 
in our observed scores $Y_i$ to be explained by the 
true but (directly) unobservable scores $\eta_i$. The measurement error $\varepsilon_i$ should be
be comparatively small (and of course on average $0$). If it is large, we are mostly measuring noise or at least not
what we want to measure. 

**How do we get to the theoretical definition of the ICC?**

If you either pull two people with the same true but unobservable score $\eta$ out of the population
or measure the same person twice and the score ($\eta$) does not change in between, we can 
**define reliability as correlation between these two measurements**:

$$Y_1 = \eta + \varepsilon_1$$
$$Y_2 = \eta + \varepsilon_2$$

$$cor(Y_1, Y_2) = cor(\eta + \varepsilon_1, \eta + \varepsilon_2) = \frac{Cov(\eta + \varepsilon_1, \eta + \varepsilon_2)}{\sigma_{Y_1}\sigma_{Y_2}}  =$$

If we use 

- the [properties of the covariance](https://en.wikipedia.org/wiki/Covariance#Properties), 
- the fact that the true score $\eta$ and the errors $\varepsilon_i$ are independent, and 
- the fact that the errors $\varepsilon_1$ and $\varepsilon_2$ are independent, we get:

$$\frac{Cov(\eta, \eta) + Cov(\eta, \varepsilon_2) + Cov(\varepsilon_1, \eta) + Cov(\varepsilon_1, \varepsilon_2)}{\sigma_{Y_1} \sigma_{Y_2}}  =$$
$$\frac{\sigma_{\eta}^2 + 0 + 0 + 0}{\sigma_{Y_1} \sigma_{Y_2}}$$

Since $\eta$ is a random variable (we draw a person randomly from the population),
it is well defined to talk about the variance of $\eta$ (i.e., $\sigma_{\eta}^2$).
I think this aspect may not come across in the book quite so clearly.

Furthermore, it does not matter if I call the measurement $Y_1$, $Y_2$ or more general
$Y$, since they have the same variance and true score: 

$$\sigma_{Y} = \sigma_{Y_1} = \sigma_{Y_2}$$

Hence, it follows that:

$$cor(Y_1, Y_2) = \frac{\sigma_{\eta}^2}{\sigma_{Y_1}^2} = \frac{\sigma_{\eta}^2}{\sigma_{Y}^2} =  \frac{\sigma_{\eta}^2}{\sigma_{Y}^2} = \frac{\sigma_{\eta}^2}{\sigma_{\eta}^2 + \sigma_{\varepsilon}^2}$$

This is the **intraclass correlation coefficient (ICC)**. 
It is (as seen in the formula above) the proportion of the
true score variability to the total variability. 
It ranges from 0 and 1 (think about why!).

A little manipulation to improve understanding:

Let's look again at the term for the ICC above and divide the numerator and the denominator by
$\sigma_{\eta}^2$, which we can do, since it is a positive number:

$$\frac{\sigma_{\eta}^2}{\sigma_{\eta}^2 + \sigma_{\varepsilon}^2} = \frac{1}{1 + \frac{\sigma_{\varepsilon}^2}{\sigma_{\eta}^2}}$$

We could call the term $\frac{\sigma_{\varepsilon}^2}{\sigma_{\eta}^2}$ the noise-to-signal ratio.
The higher this ratio, the lower the ICC. The lower the ratio, the higher the ICC.

- If you increase the noise (measurement error $\sigma_{\varepsilon}^2$) for fixed 
  true score variability $\sigma_{\eta}^2$, the ICC decreases, because the denominator
  increases.
- If you increase the true score variability $\sigma_{\eta}^2$ for fixed noise $\sigma_{\varepsilon}^2$, 
  the ICC increases, since the denominator decreases.

Btw, we could also divide by $\sigma_{\varepsilon}^2$ and get the signal-to-noise ratio.

At first glance, the following statement seems wrong:

In a very **homogeneous population** (patients have very similar scores/measurements),
the **ICC might be very low**. The reason is that the patient variability $\sigma_{\eta}^2$ is low
and you probably have some measurement error $\sigma_{\varepsilon}^2$.
Hence, if you look at the formula, ICC must be low (for a given measurement error).

On the other hand, if you have a very **heterogeneous population** (patients have rather different 
scores/measurements), the **ICC might be very high**.
The reason is that the patient variability $\sigma_{\eta}^2$ is high and you probably 
have some measurement error $\sigma_{\varepsilon}^2$.

**What matters is the ratio of the two**, as can be seen from the formula above.

Let's try to calculate the ICC for our data using a statistical model. There are a couple of different
R packages to do this. We start by using the `irr` package.

```{r}
library(irr)
irr::icc(as.matrix(df[, c("ROMas.Peter", "ROMas.Mary")]), 
    model = "oneway", type = "consistency")
cor.test(df$ROMas.Peter, df$ROMas.Mary, method = "pearson")
```

We get the result: $ICC(1) = 0.851$, which is identical to the correlation coefficient
because there is no systematic difference between Peter and Mary.

Since **we are forward looking and modern regression model experts**, 
we would like to see if we can get (approximately) the same result using the Bayesian 
framework. 

Below is the model structure (see [exercise 3](#exercise3)). 

\[
\begin{array}{rcl}
ROM_i &\sim& N(\mu_i, \sigma_{\varepsilon}) \\
\mu_i &=& \alpha[ID] \\
\alpha[ID] &\sim& \text{Normal}(\mu_{\alpha}, \sigma_{\alpha}) \\
\mu_{\alpha} &\sim& \text{Normal}(66, 20) \\
\sigma_{\alpha} &\sim& \text{Uniform}(0,20) \\
\sigma_{\varepsilon} &\sim& \text{Uniform}(0,20)
\end{array}
\]

**Model details:**

- $ROM_i$ is the observed $ROM$-score for patient $i$. Currently, we have chosen the *ROMs* to 
  come from a normal distribution. As we have seen above, this might not be adequate, 
  since the marginal distributions look skewed.
  Every patient has two observations (one each from Mary and Peter). 
  So, for instance $i=1,2$ could be patient $ID=1$.
- $\mu_i$ is the expected value of the observed score for patient $ID$.
- $\sigma_{\varepsilon}$ is the standard deviation of the measurement error.
- $\alpha[ID]$ is the patient-specific intercept ($=\eta_i$). 
  Since every patient is allowed to have a different intercept, 
  we have a **random intercepts model**. 
  The intercepts are drawn from a normal distribution (hence, *random*)
- $\mu_{\alpha}$ is the mean of the prior for the patient-specific intercepts.
  This is the overall mean of the scores.
- $\sigma_{\alpha}$ is the standard deviation of the patient-specific intercepts.
  **This is the patient variability**! $\sigma_{\alpha}$ controls how much 
  the scores of the patients may vary.
  In the extreme cases $\sigma_{\alpha} \rightarrow 0$ and $\sigma_{\alpha} \rightarrow \infty$,
  we have *complete-pooling* (all $\mu_i$ are identical) and *no-pooling* 
  (all $\mu_i$ are different and there is no shared information), respectively.
  We are using **partial pooling** here.
- The prior distributions express (as always) our prior beliefs about the parameters.

The **ICC** is then calculated as the ratio of the between-patient variance and
the total variance: 

  $$\frac{\sigma_{\alpha}^2}{\sigma_{\alpha}^2 + \sigma_{\varepsilon}^2}$$

We did not even notice it, but this was our first **multilevel regression model**.
It is multilevel due to the extra layer of patient-specific intercepts.
The **observations** are obviously **clustered within patients**, since observations
from the **same patient** are **more similar than observations from different patients**.
If one would run a normal linear regression model, one would ignore this clustering
and the assumption of independent error terms would be violated.

This time we fire up the `rethinking` package and use the `ulam` function 
to fit the model.
This uses Markov Chain Monte Carlo (MCMC) to sample from the posterior 
distribution of the parameters.

- The `chains` argument specifies how many chains we want to run.
  A *chain* is a sequence of points in a space with as many dimensions as there 
  are parameters in the model. It jumps from one point to the next in this parameter 
  space and in doing so, visits the points of the posterior approximately in the correct 
  frequency. [Here](https://blog.revolutionanalytics.com/2013/09/an-animated-peek-into-the-workings-of-bayesian-statistics.html) 
  is an excellent visualization.

- The `cores` argument specifies how many CPU cores we want to use. 
  For larger jobs, one can try to parallelize
  the chains, which saves some time.

```{r}
library(rethinking)
library(tictoc)

df_long <- df %>% 
  mutate(ID = row_number()) %>%
  dplyr::select(ID,ROMas.Peter, ROMas.Mary) %>% 
  pivot_longer(cols = c(ROMas.Peter, ROMas.Mary), 
               names_to = "Rater", values_to = "ROM") %>% 
  mutate(Rater = factor(Rater))

tic()
m5.1 <- ulam(
  alist(
    # Likelihood
    ROM ~ dnorm(mu, sigma),
    
    # Patient-specific intercepts (random effects)
    mu <- a[ID],  
    a[ID] ~ dnorm(mu_a, sigma_ID),  # Hierarchical structure for patients
    
    # Priors for hyperparameters
    mu_a ~ dnorm(66, 20),  # Population-level mean
    sigma_ID ~ dunif(0,20),  # Between-patient standard deviation
    sigma ~ dunif(0,20)  # Residual standard deviation
  ), 
  data = df_long, 
  chains = 8, cores = 4
)
toc() # 7s

precis(m5.1, depth = 2)

post <- extract.samples(m5.1)
var_patients <- mean(post$sigma_ID^2)  # Between-patient variance
var_residual <- mean(post$sigma^2)     # Residual variance
var_patients / (var_patients + var_residual) # ICC
# 0.846
# not too bad; very close to the result from the irr package
```

In the output from `precis(m5.1, depth = 2)` above we see 

- all 50 intercept estimates for each patient: `a[ID]` 
- `mu_a`is the overall intercept.
- `sigma_ID` is the **patient variability**.
- `sigma` is the **residual variability**.

We have just squared the sigmas to get the variances. 

Remember: In the background of all ICCs, there is always a statistical model to predict
the outcome. Depending on the predictors, we get different models and 
probably different ICCs.

Let's jump back into the Frequentist world and see if we can get the same result.

We can also estimate a **random intercept model** with the `lme4` package using
the command `lmer`in the Frequentist framework. No priors:

```{r}
library(lme4)
m5.2 <- lmer(ROM ~ (1|ID), data = df_long)
summary(m5.2)
print(VarCorr(m5.2), comp = "Variance")

# ICC = 
270.99 / (270.99 + 47.35) # 
# 0.8512597
# -> exactly the same result as the irr package
```

The expression `Formula: ROM ~ (1 | ID)` specifies that we want to fit a model with 
a random intercept. This means that every patient (ID) gets its own intercept
which is drawn from a normal distribution.

So far, we have only looked at the general **$ICC$** (ICC1 in the `psych`output) 
(see also page 106 in the book). 

There, we have not yet explicitely considered a bias (=systematic difference 
between the raters) that the raters could have. In the book,
they introduce a bias of 5 degrees (Mary measures 5 degrees more than Peter on average).

The **model für ICC 2 and 3** in the `psych` output explicitely considers this
(systematic) difference that could occur between the raters. 
This results in an extra term in the denominator
of the ICC, an additional variance component.

From the **same statistical model** (!) we take the variance components to calculate:

$$ICC_{agreement} = \frac{\sigma_{\alpha}^2}{\sigma_{\alpha}^2 + \mathbf{\sigma_{rater}^2} + \sigma_{\varepsilon}^2}$$

where $\sigma_{rater}^2$ is the variance due to systematic rater differences.

$$ICC_{consistency} = \frac{\sigma_{\alpha}^2}{\sigma_{\alpha}^2 + \sigma_{\varepsilon}^2}$$

**We will now introduce the 5 degree bias** and use our Bayesian 
framework to estimate the ICC.
By introducing a bias, we should see
a lower ICC. Note, that the prediction quality of Mary's scores
given Peter's scores should not change, since we would only shift Mary's scores
down by 5 degrees, which would not disturb the linear regression model. We can
always shift the points to where we want them to be. We do that for instance
when we scale or standardize the data.

Admitted, the Bayesian version in this case takes longer and is more complex.
The advantage is still that it's fully probabilistic and one could work with 
detailed prior information, especially for smaller smaple sizes.

Anyhow, let's try to give the model equations considering
the introduced bias. This is the model **for both** 
$ICC_{agreement}$ and $ICC_{consistency}$!

\[
\begin{array}{rcl}
ROM_i &\sim& N(\mu_i, \sigma_{\varepsilon}) \\
\mu_i &=& \alpha[ID] + \beta[Rater] \\
\alpha[ID] &\sim& \text{Normal}(\mu_{\alpha}, \sigma_{\alpha}) \\
\beta[Rater] &\sim& \text{Normal}(0, \sigma_{\beta}) \\
\mu_{\alpha} &\sim& \text{Normal}(66, 20) \\
\sigma_{\alpha} &\sim& \text{Exp}(0.5) \\
\sigma_{\beta} &\sim& \text{Exp}(1) \\
\sigma_{\varepsilon} &\sim& \text{Exp}(1)
\end{array}
\]

As you can see, $\mu_i$ now consists of the patient-specific intercept $\alpha[ID]$
(everyone of the 50 patients gets one)
and the rater-specific effect $\beta[Rater]$ (Mary and Peter get one). 
So, in total, we have **three sources of variability**: 

- the patient variability $\sigma_{\alpha}$, 
- the rater variability $\sigma_{\beta}$,
- and the residual variability $\sigma_{\varepsilon}$.

Note, that if Peter measures each of the 50 patients twice, 
the systematic difference between Peter's measurements would 
be zero. Of course, one could be creative and think of 
a learning effect or something.

Draw the model structure as [exercise 5](#exercise5).

```{r}
library(rethinking)
library(conflicted)
conflicts_prefer(posterior::sd)
df_long_bias <- df_long %>%
  mutate(ROM = ROM + ifelse(Rater == "ROMas.Mary", 5, 0))
head(df_long_bias)

set.seed(123)
m5.2 <- ulam(
  alist(
    # Likelihood
    ROM ~ dnorm(mu, sigma_eps),
    
    # Model for mean ROM with patient and rater effects
    mu <- alpha[ID] + beta[Rater],
    
    # Patient-specific random effects
    alpha[ID] ~ dnorm(mu_alpha, sigma_alpha),
    
    # Rater effect (Peter/Mary)
    beta[Rater] ~ dnorm(0, sigma_beta),
    
    # Priors for hyperparameters
    mu_alpha ~ dnorm(66, 10),  # Population mean ROM
    sigma_alpha ~ dexp(0.5),  # Between-patient SD (less aggressive shrinkage)
    sigma_beta ~ dexp(1),   # Rater SD (better regularization)
    sigma_eps ~ dexp(1)     # Residual SD (prevents over-shrinkage)
  ), 
  data = df_long_bias, 
  chains = 8, cores = 4
)

precis(m5.2, depth = 2)
precis(m5.2)

# check systematic difference for rater in posterior
post <- extract.samples(m5.2)
mean(post$beta[,1] - post$beta[,2])

# ICC agreement:
post <- extract.samples(m5.2)
(var_patients <- mean(post$sigma_alpha^2))  # Between-patient variance
(var_raters <- mean(post$sigma_beta^2))     # Rater variance
(var_residual <- mean(post$sigma_eps^2))    # Residual variance

# ICC_agreement = 
var_patients / (var_patients + var_raters + var_residual)
# 0.8033613 (sigma_alpha ~ dexp(1))
# 0.83 (sigma_alpha ~ dexp(0.5))

# ICC (Single_fixed_raters) = ICC3 in psych output = 
var_patients / (var_patients + var_residual)
# 0.8415256
```

It should be noted that this ICC is very sensitive to the choice of the prior.
If you choose too agressive priors for the standard deviations $\sigma_{\alpha}, 
\sigma_{\beta}, \sigma_{\varepsilon}$, you will get a too low ICC.

I have played around a little with the parameters in the exponential priors 
to get the desired result which compares nicely to the two alternative methods below:
using the `psych` package and with the `lmer`
package. Both use a Frequentist random intercept model in the background.
Using a package like `psych` gives a rather convenient interface to 
elicit the ICC. It can never hurt to peek behind the curtain a little bit though.

**`psych` package** (Frequentist):
```{r}
library(psych)
library(conflicted)
# needs wide format
#conflicts_prefer(dplyr::select)
df_wide <- df_long_bias %>%
  pivot_wider(names_from = Rater, values_from = ROM)
df_wide_values <- df_wide %>% dplyr::select(-ID)
psych::ICC(df_wide_values) # ICC1 = 0.83
```

**`lmer` package**:
```{r}
# _lmer------
m5.3 <- lmer(ROM ~ (1 | ID) + (1 | Rater), data = df_long_bias)
summary(m5.3)
print(VarCorr(m5.3), comp = "Variance")
# Groups   Name        Variance
# ID       (Intercept) 270.882 
# Rater    (Intercept)   6.193 
# Residual              47.557 


# ICC (Single_random_raters) = ICC2 in psych output
270.882 / (270.882 + 6.193 + 47.557) # 
# 0.8344279

# ICC (Single_fixed_raters) = ICC3 in psych output
270.882 / (270.882 + 47.557) #
# 0.85
```

There are basically **two different random effects**:

- **Nested random effects**: The patients (respectively their $ROMs$) would cluster within the raters.
  This would mean, a patient would only be measured by Peter twice or by Mary twice.
  In that case, the $ROMs$ would cluster within the patients (because of the repeated measurement)
  and all patients measured by (for instance) Mary would cluster together (and be potentially more similar).
  In this case all patients are measured by Peter and Mary, so we do not have nested random effects.
  In `lmer` this would be specified as `ROM ~ (1|Rater/ID)`.
- **Crossed random effects**: 
  Not nested.
  In R this would be specified as `ROM ~ (1|ID) + (1|Rater)`.
  
### Explanation of ICCs in the `psych` output

If you want to know all the details, refer to [Shrout and Fleiss (1979)](https://d1wqtxts1xzle7.cloudfront.net/50483847/syarat_reliabilitas_icc-libre.pdf?1479841049=&response-content-disposition=inline%3B+filename%3DIntraclass_Correlations_Uses_in_Assessin.pdf&Expires=1740684631&Signature=hHiFbcQD3PDVIyWDJ-bUhcm3WtsK19YhHm6FKtnafNdqsm9NhR6cr9lbCf~gVV5SYG1XlTwLlcfJkQ9Z-ahjmmNV893aWi5plo~yL4oZBEjrmFa9WCd7k6vzFTkri1Xbgfh~GyPARWXBtqABytovtL-RD1420Kw9qk150nw3-kUWcuvRiIc~r0y65XQaXf-V9mm~uXRFdUqec4Vs-Bwh~IrJfHWQASGgp8wZjzh2130MCP3-iaorxNn~79c~nm2f1aIl5WRqRXB6EIy8HlrNFpxNSt1pgTPZoZadEECM4qH395KLY5ijUnhoCDT9AmcOplPnFiC5t8dKW-n25ziofQ__&Key-Pair-Id=APKAJLOHF5GGSLRBV4ZA).
The help-function `?psych::ICC` contains a relatively good and much shorter explanation.
The variance formulae given in the help-file are probably somewhat confusing.
We try to stick to the notation of the book.

Let's talk about the first three **ICCs in the `psych` output**:  

- **Single_raters_absolute ICC1:**
  According to the help file: "Each target [i.e. patient] is rated by a different 
  judge [i.e. rater] and the judges are selected at random." So, variability due to raters
  is implied and cannot be disentangled. 
  This is formally not our situation, since we have only two raters and
  50 patients. But for this case, we do not care who measures, since we do not
  model it, hence, we cannot know if there are systematic differences between
  the raters. There might as well be 50 raters doing their thing, or just 2 as in our case. 
  This is the ICC we have calculated above; the overall ICC 
  in the book and it is based on the following model:
  
  $$Y_{ij} = \eta_i + \varepsilon_i$$
  where $i \in {1,...,50}$ is the *patient* and $j \in {1,2}$ is the *measurement* 
  ($=50*2=100$ rows in long format).
  Note that we do not mention a rater here, since we do not care who took the
  measurement. It is not part of the model. The ICC is then calculated as:

  $$ICC = \frac{\sigma_{\eta}^2}{\sigma_{\eta}^2 + \sigma_{\varepsilon}^2}$$
  
  whereas we could get the variance components from either the posterior in the Bayesian
  setting or from the `lmer` output in the Frequentist setting.

- **Single_random_raters ICC2:** 
  ICC2 ($=ICC_{agreement}$ in the book) and ICC3 ($=ICC_{consistency}$ in the book) 
  are based on the **same (!)** statistical model. The only difference is
  that ICC2 assumes that the (in our case) 2 raters are randomly selected from a larger pool of raters,
  hence, the rater variability must be explicitely considered and yields a potentially smaller
  value for the ICC. Compared to ICC1, we have repeated measurements from the same raters
  in 50 patients. That's why we can model their bias. One observation would not be enough.
  The help file says: "A random sample of k judges rate each target. 
  The measure is one of absolute agreement in the ratings."
  A random sample of k (2 in our case) judges means that we cannot rule out the variability
  due to raters (you get a variety of them and their biases are different).
  
  $$Y_{ij} = \eta_i + \beta_j + \varepsilon_i$$
  where $i \in {1,...,50}$ is the *patient*, $j \in {1,2}$ is the **rater** 
  (doing one measurement in each patient). The ICC is then calculated as:
  
  $$ICC_{agreement} = \frac{\sigma_{\eta}^2}{\sigma_{\eta}^2 +\mathbf{\sigma_{rater}^2} + \sigma_{\varepsilon}^2}$$

- **Single_fixed_raters ICC3:**
  ICC3 ($=ICC_{consistency}$ in the book) is based on the **same model as ICC2**, 
  but assumes that the raters are fixed. 
  This means that **the raters are the same for all patients** in the future study. 
  So, we have considered the rater variability in the model 
  (which was possible due to the repeated measurements from the same raters in 50 patients), 
  but do not care since Mary and Peter will be the people doing the future
  measurements, not other therapists. If you fix a random variable 
  (*raters* in this case), variance is zero. The help file says: 
  "A fixed set of k judges rate each target. 
  There is no generalization to a larger population of judges."
  The ICC is then calculated as:

  $$ICC_{consistency} = \frac{\sigma_{\eta}^2}{\sigma_{\eta}^2 + \sigma_{\varepsilon}^2}$$

  **Important**: ICC1 and ICC3 are **not** identical, since ICC1 does not consider the rater variability
  in the model. They are based on *different* statistical models.
  ICC2 and ICC3 are based on the *same* model.

If there is no systematic difference between raters, all 3 ICCs and the Pearson 
correlation (r) are the same (see Figure 5.3 in the book).

$ICC_{consistency}$ vs. $ICC_{agreement}$:
The latter is used, when we need Peter and Mary to concur in their measurements.
Patients coming to Peters practice will get the same (or very similar) "diagnosis" (ROM-value)
from Mary. When there is systematic difference (line is shifted downwards in Figure 5.3), 
this cannot be guaranteed.
If we only need Peter and Mary to *rank* the patients in the same order,
we can use $ICC_{consistency}$.

- **Average_raters_absolute ICC1k:**
  According to the help file "ICC1k, ICC2k, ICC3k reflect the means of k raters."
  In general, if you take the mean, you get an estimate with lower variance (central limit theorem). 
  In the book on page 109, we find the derivation of the ICC for the average of k raters.
  We do not need a new statistical model, although we take the sum (or equivalently, the mean) of the
  measurements of Peter and Mary, which seems a bit unintuitive at first. 

  If we go back to [the definition of the ICC above](#ICC), it follows that:

  $$Y_{1} + Y_{2} = 2 \eta + \varepsilon_1 + \varepsilon_2$$ 
  for the same person with true $\eta$. Since ther errors $\varepsilon_i$ are 
  independent and the true score $\eta$ is also independent from the errors, 
  it follows that:

  $$\mathbb{Var}(Y_1 + Y_2) = \mathbb{Var}(2 \eta + \varepsilon_1 + \varepsilon_2)=$$
  $$ \mathbb{Var}(2 \eta) + \mathbb{Var}(\varepsilon_1 + \varepsilon_2)=$$ 
  $$4 \mathbb{Var}(\eta) + 2 \mathbb{Var}(\varepsilon)=$$

  If we divide by 4, we get:
  $$\frac{1}{4}\mathbb{Var}(Y_1 + Y_2) = \mathbb{Var}(\frac{Y_1 + Y_2}{2}) = 
  \mathbb{Var}(\eta) + \frac{\mathbb{Var}(\varepsilon)}{2} = 
  \sigma_{\eta}^2 + \frac{\sigma_{\varepsilon}^2}{2}$$

  Hence, the ICC for the average of 2 raters is ($k=2$):

  $$ICC1k = \frac{\sigma_{\eta}^2}{\sigma_{\eta}^2 + \frac{\sigma_{\varepsilon}^2}{2}}$$

  Again, the model does not care about the raters, it just knows that there are 
  two measurements. A potential bias between raters is not considered.

  It is also interesting to note that we do not fit a new model to get the ICC1k.
  We could not compare the first average (of 2 measurements) to the second average 
  (of 2 measurements), since we only have 2 measurements per patient.

```{r}
m <- lmer(ROM ~ (1|ID), data = df_long_bias)
summary(m)
print(VarCorr(m), comp = "Variance")

# ICC1k = 
267.79 / (267.79 + 53.75/2)
```

This is identical to the ICC1k in the `psych` output.

ICC1k would approach $1$ if $k$ goes to infinity. 
**This makes sense**, since the *true* measurement is *defined* as the average of 
infinite measurements (of either infinitely many different raters or infinitely many 
measurements from the same rater). 
Also it makes intuitive sense, that 
reliability increases or equivalently the unexplained variance decreases
since we draw repeatedly from the same distribution ($\eta$ and an error term $\varepsilon$)
and the central limit theorem guarantees that we can estimate $\eta$ as precicely 
as we want (variance of the mean goes to zero) if we only draw often enough.

- **Average_raters_random ICC2/3k:**
  ICC2k and ICC3k are based on the same model as ICC2 and ICC3, respectively.
  We just need to divide the respective residual variance by $k$ (in our case 2) to get 
  the ICC for the average of $k$ raters.

```{r}
m5.3 <- lmer(ROM ~ (1 | ID) + (1 | Rater), data = df_long_bias)
summary(m5.3)
print(VarCorr(m5.3), comp = "Variance")
# Groups   Name        Variance
# ID       (Intercept) 270.882 
# Rater    (Intercept)   6.193 
# Residual              47.557 

# ICC2k =
270.882 / (270.882 + (6.193 + 47.557) / 2)

# ICC3k =
270.882 / (270.882 + 47.557 / 2)
```

These values are identical to the ICC2k and ICC3k in the `psych` output.


### Summary Peter and Mary, with and without bias (ICC1-3)

Below, we summarize the results for the ICCs (calculated with `psych`)
for the unbiased and biased case (Mary measures on average 5 degrees more than peter).

```{r}
library(pacman)
p_load(conflicted, tidyverse, flextable)

# Ensure select() from dplyr is used
#conflicted::conflicts_prefer("select", "dplyr")

# Unbiased ICC Calculation
df_wide_unbiased <- df_long %>%
  pivot_wider(names_from = Rater, values_from = ROM)
df_wide_values_unbiased <- df_wide_unbiased %>% dplyr::select(-ID)
icc_results_unbiased <- psych::ICC(df_wide_values_unbiased)

# Extract relevant ICC values
icc_unbiased_df <- icc_results_unbiased$results %>%
  dplyr::select(type, ICC) %>%
  rename(`Unbiased ICC` = ICC)

# Biased ICC Calculation
df_wide_biased <- df_long_bias %>%
  pivot_wider(names_from = Rater, values_from = ROM)
df_wide_values_biased <- df_wide_biased %>% dplyr::select(-ID)
icc_results_biased <- psych::ICC(df_wide_values_biased)

# Extract relevant ICC values
icc_biased_df <- icc_results_biased$results %>%
  dplyr::select(type, ICC) %>%
  dplyr::rename(`Biased ICC` = ICC)

icc_merged_df <- left_join(icc_unbiased_df, 
                           icc_biased_df, 
                           by = "type") %>%
  slice(1:3)

ft <- flextable(icc_merged_df) %>%
  flextable::set_header_labels(type = "ICC Type") %>%
  flextable::set_caption("Intraclass Correlation Coefficients - Unbiased vs. Biased") %>%
  flextable::set_table_properties(width = .5, layout = "autofit")
ft
```

The **left column** shows that the ICCs are identical for the **unbiased case**.
Specifically, ICC2 (=$ICC_{agreement}$ in the book) and 
ICC3 ($ICC_{consistency}$ in the book) are based on the same model which explicitely 
considers a potential bias between the raters. Since there is none,
the ICCs are the same.

In the **biased case** (right column), there *is* as systematic difference between Mary and Peter - 
5 degrees in our case.

**ICC1** does not care about it and shows a somewhat lower value compared to 
before ($0.833$). The reason is because the agreement line ($y=x$) in a plot with 
Mary on Y and Peter on X is shifted upwards by 5 degrees.
If you would introduce a bias of 15 degrees, the ICC would
be even lower ($ICC1 = 0.61$, $ICC2 = 0.65$, [exercise 1](#exercise1)). 
The unbiased column would of course stay the same.

**ICC2** now considers the bias of 5 degrees. The model knows about the shift.
If we compare the variance components of ICC1 and ICC2, we see:
  
```{r, echo=FALSE}
print("Model for ICC1: ROM ~ (1 | ID)")
m_ICC1 <- lmer(ROM ~ (1 | ID), data = df_long_bias)
print(VarCorr(m_ICC1), comp = "Variance")

print("Model for ICC2 (and 3): ROM ~ (1 | ID) + (1 | Rater)")
m_ICC2 <- lmer(ROM ~ (1 | ID) + (1 | Rater), data = df_long_bias)
print(VarCorr(m_ICC2), comp = "Variance")
```

The residual variance is smaller in the model with the rater effect.
The model explains the data better, since it knows about the bias (see [exercise 7](#excercise7)).

Look at the $\sigma_{\varepsilon}^2$ of the two models, they add up:

$$\sigma_{\varepsilon, ICC1}^2 = \sigma_{\varepsilon, ICC23}^2 + \sigma_{Rater}^2$$
$$53.75 = 47.557 + 6.193$$

Hence, we just split up the error differently by considering the bias.
The patient variability (`ID Variance` in the output) is slightly higher: 
$270.882$ compared to $267.79$ before.
In the first model, patient variability was conflated with rater variation/systematic difference
because rater effects were not explicitly modeled. For this reason, 
`ID Variance` increases slightly.
It is a rather small increase, so ICC1 and ICC2 are not that different.

For a bias of 15 degrees, the additivity of the variances remains.
The patient variability increases from $223.89$ (ICC1) to $270.882$ (ICC2),
hence the difference in ICCs is larger ($ICC1=0.613$ vs. $ICC2=0.657$).

**ICC3** considers the bias in the model but does not include it in the measurement 
error since the raters are fixed.


### Difference between correlation and ICC

If we do not introduce a bias in the data, the correlation coefficient
is the same as the ICC (as seen above). On page 110, Figure 5.3, the authors show nicely 
what the difference is between the correlation coefficient and the ICC. 
It is also shown how $ICC_{agreement}$ and $ICC_{consistency}$ change with the bias.

$ICC_{consistency}$ stays $1$ if bias is introduced (assuming a hypothetical 
perfect agreement before). Peter and Mary still rank the patients
in the same order.

Correlation $r$ is always 1, no matter at what slope ($\ne 0$) the line is.
It measures the strength and direction of the *linear* relationship between two variables.

$ICC_{agreement}$ changes as soon as you depart from the 45 degree line
with respect to slope or 
shift the line up or down (i.e., introduce a bias).

This is a good point in time to think for a moment about the type of measurement
for agreement with respect to **costs**. The ICC below weights each measurement
equally, although one larger outlier (large discrepancy between Peter and Mary)
may influence the ICC notably ([exercise 8](#exercise8)). 
One could possibly think of a situation where 
overall agreement measure like the ICC is not adequate since, 
for instance, exceeding a certain
difference threshold could decide between life and death. 

### Bad news about the ICC? {#badnews}

Let's try to expand our intuitive understanding of what an ICC of 0.8 or so means.
For simplicity, we take the overall ICC, which is equal to the correlation coefficient,
if there is no bias present.

The following example demonstrates the meaning of the Test-Retest reliability of 
the Hospital Anxiety and Depression Scale - Anxiety subscale (HADS-A). We could take 
all kinds of other scores where ICC values and a 
[minimally clinically important difference (MCID)](https://en.wikipedia.org/wiki/Minimal_important_difference)
is given. Briefly, the MCID is the smallest change in a score that is considered
important to the patient. For example, a 5% change in BMI is (in some populations)
considered meaningful.

Specifically, we:

- Simulate two correlated measurements at two time points (TP1 and TP2) 
  to get predetermined ICC ($=\rho$).
- Calculate the Intraclass Correlation Coefficient (ICC).
- Compare the Minimally Clinically Important Difference (MCID) to prediction intervals.
- Visualize how often a clinically meaningful change of 1.68 points is 
  detected, even if no real change has occurred.

The code can be found [here](https://github.com/jdegenfellner/ICC_MCID/blob/main/ICC_MCID.R)
and in the github repo of the script. In the code, the sources for the ICC and MCID are
cited.

```{r}
# ICC and Test-Retest-Reliability

library(pacman)
p_load(tidyverse, lme4, conflicted, psych, MASS)

# MCID Minimal Clinically Important Difference---------

# HADS score---------
# The Hospital Anxiety and Depression Scale 

# Test-Retest-Reliability:
# https://doi.org/10.1016/S1361-9004(02)00029-8

# Use the numbers from here (Table 1):
# https://www.sciencedirect.com/science/article/abs/pii/S1361900402000298
# for demonstration purposes.

# Minimal Clinically Important Difference (MCID) for HADS-A:
# https://pmc.ncbi.nlm.nih.gov/articles/PMC2459149/
# Not exactly the same population as shift workers, but suffices for demonstration purposes.
# MCID HADS anxiety score and 1.68 (1.48–1.87)

# Simplification: Score is deemed to be continuous.

# Create 2 correlated measurements:
# Use HADS-A Anxiety subscale
# Use n=100, instead of n=24 for more stability

sigma1 <- 3.93  # Standard deviation of variable 1
sigma2 <- 3.52  # Standard deviation of variable 2
rho <- 0.82    # Correlation
cov_matrix <- matrix(c(sigma1^2, rho * sigma1 * sigma2,
                       rho * sigma1 * sigma2, sigma2^2), nrow = 2)
n <- 100

set.seed(188)  # For reproducibility
samples <- mvrnorm(n = n, mu = c(7.92, 7.83), Sigma = cov_matrix)
df <- as.data.frame(samples)
colnames(df) <- c("TP1", "TP2")

plot(df, main = "Scatterplot of Multivariate Normal Samples",
     xlab = "TP1", ylab = "TP2", pch = 19, col = rgb(0, 0, 1, alpha = 0.5))

# https://en.wikipedia.org/wiki/Intraclass_correlation#Modern_ICC_definitions:_simpler_formula_but_positive_bias
# ICC should be the correlation within the group (i.e. patient)

cor(df$TP1, df$TP2, method = "pearson") # ~0.8-0.9
ICC(df) # 0.82

# check manually
df_mod <- data.frame(id = 1:n, df$TP1, df$TP2)
names(df_mod) <- c("id", "TP1", "TP2")
df_mod_long <- df_mod %>% pivot_longer(cols = c(TP1, TP2), 
                                       names_to = "time_point", 
                                       values_to = "score")
mod <- lme4::lmer(score ~ time_point + (1|id), data = df_mod_long)
variance_df <- as.data.frame(summary(mod)$varcor)
# ICC=
variance_df$sdcor[1]^2 / (variance_df$sdcor[1]^2 + variance_df$sdcor[2]^2) # ~0.9
# cor and ICC are very very similar, should actually be identical.

#data.frame(TP1 = df$TP1, TP2 = df$TP2) %>% 
#  ggplot(aes(x = TP1, y = TP2)) +
#  geom_point() +
#  xlab("Measurement of patients at time point 1") +
#  ylab("Measurement of the same patients at time point 2")

# Range for HADS-A should be 0-21 
# (according to "The Hospital Anxiety and Depression Scale, Zigmond & Snaith, 1983")

df <- data.frame(TP1 = df$TP1, TP2 = df$TP2) %>% 
  dplyr::filter(TP1 >= 0) %>%
  dplyr::filter(TP2 >= 0) %>% # negative not possible
  dplyr::filter(TP1 <= 21) %>%
  dplyr::filter(TP2 <= 21) # max. score is 21
df

mod <- lm(TP2 ~ TP1, data = df)
pred <- predict(mod, df, interval = "prediction")

# How wide are the prediction intervals for a patient?
as.data.frame(pred) %>% mutate(width_prediction_interval = upr - lwr) # width of the prediction interval 8-10 points!

# Example:
predict(mod, newdata = data.frame(TP1 = 10), interval = "prediction") # 95% prediction interval for a patient with a score of 10 at time point 1.
(13.56369 - 5.226282)/1.68

# Prediction interval width is ~5 times our minimally clinically important 
# change of 1.68 for HADS-A.

df %>% 
  ggplot(aes(x = TP1, y = TP2)) +
  # Color points conditionally
  geom_point(aes(color = ifelse(TP2 > TP1 + 1.68 | TP2 < TP1 - 1.68, "red", "black"))) +
  scale_color_manual(values = c("red" = "red", "black" = "black"), guide = "none") +
  geom_abline(intercept = mod$coefficients[1], slope = mod$coefficients[2]) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_ribbon(aes(ymin = pred[,2], ymax = pred[,3]), alpha = 0.2) + 
  ggtitle("HADS-A and 95% Prediction Interval for TP2") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(intercept = 0, slope = 1, color = "green", linetype = "dashed") +
  geom_abline(intercept = 1.68, slope = 1, color = "red", linetype = "dashed") +
  geom_abline(intercept = -1.68, slope = 1, color = "red", linetype = "dashed")

# How often is TP2 within the MCIC of 1.68 points?---------
df$abs_diff <- abs(df$TP1 - df$TP2)
hist(df$abs_diff)
abline(v = 1.68, col = "red")

table(df$abs_diff > 1.68)/sum(table(df$abs_diff > 1.68)) # 
# -> in ~46% of cases we detect a minimally clinically important change of 1.68 points.
# even though the underlying truth did not change.

# This result is near random guessing
```

The example shows that for the given value of ICC ($>0.8$) and MCID, the test retest reliability 
is near random guessing with respect to detecting a change, since the second prediction is 
very often outside the MCID (red dashed lines). Or in short: Even if Mary measures twice 
with this ICC value, she will detect a clinically meaningful change in almost half of the cases, 
although the underlying
truth did not change. 

One could play around a little bit with the parameters here 
(and maybe make the argument more rigorous),
but my guess would be that the main message remains stable.

What do you think? Can you find literature or logical arguments against this
simulation? I would be pleased to hear your thoughts.

What could come to the rescue of this systematic problem?
I think one must consider more reliable scales simultaneously in order
to make decicisions, specifically to determine if a change occured.

See [exercise 4](#exercise4) to try the above code with a score of your choosing.

### Standard Error of Measurement (SEM)
Above, we defined the overall ICC as:

$$ICC = \frac{\sigma_{\alpha}^2}{\sigma_{\alpha}^2 + \sigma_{\varepsilon}^2}$$

The SEM is defined as: $\sqrt{\sigma_{\varepsilon}^2}$.

This is a measure for the precision of the model. If this number is small,
we know the true but unknown score ($\eta$) very precisely. As we have seen, for the
$ICC_{agreement}$ the error term includes the rater variability $\sigma_{rater}$.

For ICC2 and 3 the model is the same and:

$$Y_{ij} = \eta_i + \beta_j + \varepsilon_i$$

For $ICC_{agreement}$, the SEM is then:
$$\sigma_{Y_i}^2 = \sigma_{\eta}^2 + \underbrace{\sigma_{\text{rater}}^2 + \sigma_{\varepsilon}^2}_{SEM_{agreement}^2}$$


For $ICC_{consistency}$, the SEM is then:
$$\sigma_{Y_i}^2 = \sigma_{\eta}^2 + \underbrace{\sigma_{\varepsilon}^2}_{SEM_{consistency}^2}$$

And since ICC 2 and 3 are based on the same model:
$$SEM_{consistency} \le SEM_{agreement}$$
(equality if there is no bias)

This yields the **first method** of getting the SEM: Estimate the 
model (as we did above) and take the square root of the respective error term
(with or without the rater variability).

One can also show that we can find the $SEM_{consistency}$ by using
the standard deviation of the differences between the two measurements
of Peter and Mary:

We verify this immediately:

$$SEM_{consistency} = \frac{SD_{difference}}{\sqrt{2}}$$

```{r}
mod <- lmer(ROM ~ (1 | ID) + (1 | Rater), data = df_long_bias)
print(VarCorr(mod))
```

$$SEM_{consistency} = \sigma_{residual} = 6.8961$$

Let's calculate the SEM from the differences between the two measurements:

```{r}
# SEM from differences
sd(df_long_bias$ROM[df_long_bias$Rater == "ROMas.Mary"] -
   df_long_bias$ROM[df_long_bias$Rater == "ROMas.Peter"]) / sqrt(2)
```

This is a perfect match! 

### Bland-Altman Plot

The Bland-Altman plot (see 5.4.2.2. in the book) is a graphical method to compare 
two measurements techniques or raters. The goal is to spot systematic differences 
between the two
methods/raters. On the x axis, we plot the average of the two measurements (proxy for the true value)
and on the y axis the difference between the two measurements. So, we could 
for instance see larger variability in points with higher average values,
indicating that Peter and Mary disagree more for higher values.

Let's quickly draw one and explain it:

```{r}
library(BlandAltmanLeh)
library(data.table)
df_long_bias <- as.data.table(df_long_bias)
# https://cran.r-project.org/web/packages/BlandAltmanLeh/BlandAltmanLeh.pdf
bland.altman.plot(df_long_bias[Rater == "ROMas.Mary", ]$ROM,
                  df_long_bias[Rater == "ROMas.Peter", ]$ROM,
                  two = 1.96,
                  mode = 1,
                  conf.int = 0.94
                  )
# "two": Lines are drawn "two" standard deviations from mean differences. 
# This defaults to 1.96 for proper 95 percent confidence interval 
# estimation but can be set to 2.0 for better agreement with e. g. 
# the Bland Altman publication. 

# "mode": if 1 then difference group1 minus group2 is used,
# if 2 then group2 minus group1 is used. Defaults to 1.

# "conf.int": confidence interval for the mean difference
# and the limits of agreement.

# The red dotted line is the mean difference $\bar{d}$ 
# between the two measurements.

# see ?bland.altman.plot
```

The *Limits of Agreement* (LoA) are derived from data and represents
the "usual" values for the difference between the two measurements. 
The $1.96$ in the formula is (as always)
a hint that we assume the difference is normally distributed. This is not 
always the case. 

The formula for the LoA is (due to the connection between the SEM and the 
SD of the differences above):

$$LoA = \bar{d} \pm 1.96 \cdot SD_{difference} = 
\bar{d} \pm 1.96 \cdot \sqrt{2} \cdot SEM_{consistency} \cdot SD_{difference}$$

Let us not get confused by the name. Just because a data point is shown
within the LoA, it does not mean that it is a good measurement. Maybe $\pm 20$ degrees
(as visible in the plot) is unsifficient agreement for practical purposes.
The LoA is a statistical measure, not a clinical one.

To remember the fact that a normal distribution is assumed for the differences
$\bar{d}$ is assumed, we draw the limits of agreement for our data:

```{r, echo=FALSE}
# ────────────────────────── Pakete ──────────────────────────
library(data.table)        # effiziente Datenmanipulation
library(BlandAltmanLeh)    # schnelle Bland-Altman-Plots
library(ggplot2)           # Grafik-Grammar

# ─────────────────── Daten vorbereiten ──────────────────────
setDT(df_long_bias)  # sicherstellen, dass es ein data.table ist

# Messreihen der beiden Rater extrahieren
x <- df_long_bias[Rater == "ROMas.Mary",  ROM]   # Rater 1
y <- df_long_bias[Rater == "ROMas.Peter", ROM]   # Rater 2

# ───────────────── Normaldichte der Differenzen ─────────────
d        <- x - y
d_bar    <- mean(d, na.rm = TRUE)        # Mittelwert der Differenzen
sd_diff  <- sd(d,   na.rm = TRUE)        # SD der Differenzen
LoA_low  <- d_bar - 1.96 * sd_diff       # unteres Limit
LoA_high <- d_bar + 1.96 * sd_diff       # oberes Limit

# Sequenz für die x-Achse (±4 SD um den Mittelwert)
x_seq <- seq(d_bar - 4*sd_diff,
             d_bar + 4*sd_diff,
             length.out = 400)

df_plot <- data.frame(
  x    = x_seq,
  dens = dnorm(x_seq, mean = d_bar, sd = sd_diff)
)

ggplot(df_plot, aes(x, dens)) +
  geom_line(linewidth = 1) +
  # Mittelwert-Linie
  geom_vline(xintercept = d_bar, linetype = "dashed") +
  # LoA-Linien
  geom_vline(xintercept = c(LoA_low, LoA_high), linetype = "dotted") +
  # Beschriftungen
  annotate("text", x = d_bar,    y = 0,
           label = sprintf("d̄ = %.2f", d_bar),
           angle = 90, vjust = -0.6, hjust = -1) +
  annotate("text", x = LoA_low,  y = 0,
           label = sprintf("LoA = %.2f", LoA_low),
           angle = 90, vjust = -0.6, hjust = -1.5) +
  annotate("text", x = LoA_high, y = 0,
           label = sprintf("LoA = %.2f", LoA_high),
           angle = 90, vjust = -0.6, hjust = -1.5) +
  labs(title = "Limits of Agreement (biased Peter and Mary, 50 ROMs)",
       x     = expression(bar(d)~"(difference)"),
       y     = "") +
  theme_minimal(base_size = 13) + 
  theme(plot.title = element_text(hjust = 0.5))

# \bar{d}=3.87 (before bias introduction, Mary was on average 1.2 lower)
```

$\bar{d}=3.87$ (before the bias introduction, Mary was on average 1.2 lower)

See also [exercise 6](#exercise6).

### Example of a Reliability Study {#example_RELStudy}

Let's try to verify some of the ICCs calculated in 
[this paper](https://www.biorxiv.org/content/10.1101/2024.06.23.600286v2.abstract).

The title of the paper is "Between-day reliability of local 
and global muscle-tendon unit assessments in female athletes 
whilst controlling for menstrual cycle phase".

It takes a while to find a study that provides the data. Kindly, the authors
provide raw data [here](https://figshare.com/articles/dataset/Test-retest_dataset_xlsx/26477206?file=48128929).

Some comments after a quick glance at the paper:

- One thing you should avoid is to write results in the methods section as
  is done in lines 152-154 of the [pdf](https://www.biorxiv.org/content/10.1101/2024.06.23.600286v2.full.pdf).
  First we describe what we did, then we show the results.

- The next thing to avoid is testing for normality. For seventeen participants,
  the power of the test might be low - see Figure 1a/b [here](https://www.nbi.dk/~petersen/Teaching/Stat2017/Power_Comparisons_of_Shapiro-Wilk_Kolmogorov-Smirn.pdf).
  
- Apparently, there was no power calculation for the ICCs or the paired $t$-test.
  As a consequence, results (for instance for the $t$-test) are to be interpreted exploratively.

- Multiple testing is not considered.
  From the first $28$ $p$-values in Table 1 and 2 of the paper, two are "significant",
  which is well within the expected number of false positives. Remember,
  under $H_0$ every $20$th test is "significant".
  Fortunately, the authors do not overinterpret the results and 
  do not mention "significance" in the results or discussion.

Look at Table 1 in the paper and consider the variable *Planar flexion MVC (N.m)*.
We want to replicate the results in the first line 
(except for $p$-values and effect size $d$).

```{r}
library(pacman)
conflicts_prefer(rethinking::logit)
p_load(
  tidyverse,
  readxl,
  psych)

# Read file from github
tmp <- tempfile(fileext = ".xlsx")
download.file("https://raw.githubusercontent.com/jdegenfellner/Script_QM3_ZHAW/main/data/Chapter_Rel_Val/2_bioxar/Test-retest_dataset.xlsx", 
              destfile = tmp, mode = "wb")
df <- suppressWarnings(suppressMessages(read_xlsx(tmp, sheet = "Sheet1", skip = 1)))
head(df)
#colnames(df)

df_1 <- df %>% dplyr::select(`Plantar flexion moment (N.m)`, `...11`)
head(df_1)
#tail(df_1)
dim(df_1)

ggplot(df_1, aes(x = `Plantar flexion moment (N.m)`, y = `...11`)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = "Scatterplot of Plantar flexion moment",
    x = "Plantar flexion moment (N.m) - Session 1",
    y = "Plantar flexion moment (N.m) - Session 2"
  ) +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
# Looks rather nice
# would be interesting what influence the largest 2 and the smallest observation have

colnames(df_1) <- c("test", "retest")

# Mean Session 1
mean(df_1$test, na.rm = TRUE)
# 124.18 check
sd(df_1$test, na.rm = TRUE)
# 15.91872 check

# Mean Session 2
mean(df_1$retest, na.rm = TRUE)
# 126.1748 check
sd(df_1$retest, na.rm = TRUE)
# 20.41306

# Change
mean(df_1$retest - df_1$test, na.rm = TRUE)
# 1.99477 check
sd(df_1$retest - df_1$test, na.rm = TRUE)
# 8.160476 check

# The p-value column is not needed, as well as the d (effect size)

# create long format for ICC calculation
df_long <- df_1 %>%
  pivot_longer(cols = everything(), names_to = "Timepoint", values_to = "value") 
head(df_long)

psych::ICC(df_1)

# In Table 1, we find the values for ICC1k: 0.95 (95% CI: 0.86-0.98)

# Standard error of measurement:
# SEM = SD_difference/sqrt(2)
sd(df_1$retest - df_1$test, na.rm = TRUE)/sqrt(2)
# 5.770328 vs. 4.2 in the paper

# In "Measurement in Medicine", p. 112, the authors warn against
# the formula the paper used to calculate the SEM:
# SEM = SD_pooled*sqrt(1-ICC)
SD_pooled <- sqrt((sd(df_1$test, na.rm = TRUE)^2 + sd(df_1$retest, na.rm = TRUE)^2)/2) # correct?
SEM <- SD_pooled*sqrt(1 - 0.95)
SEM
# 4.092977 vs. 4.2 in the paper (rounding error?)

# Formula 5.7 in the book:
# sigma_y is the total variance
# lets get the variance components with lmer:
dim(df_long)
df_long$ID<- rep(1:17, each=2)
mod <- lmer(value ~ Timepoint + (1|ID), data = df_long)
summary(mod)
# check ICC1k:
301.8/(301.8 + 33.3/2)
# 0.9477155 check

# -> sigma_y = SD_pooled = 
sqrt(301.8 + 33.3) # = 18.30574

# or this?
sqrt(301.8 + 33.3/2) # 17.84517
# SEM = 
sqrt(301.8 + 33.3/2)*sqrt(1 - 0.95) # 3.990301

# -> SEM was not exactly replicable

# MCD, Minimally Detectable Difference:
# (also known as Smallest Detectable Change, SDC)
# MCD = 
1.96 * 4.092977 * sqrt(2)
# 11.34515 vs 11.6 in the paper, rather close.

1.96 * 4.2 * sqrt(2)
# 11.64181

# Coefficient of Variation:
# The CV was calculated for each participant as (within-subject SD / mean) * 100, 
# with the mean of values from all participants used as the test-
# retest CV (e.g., [24]).

df_1$mean_i <- rowMeans(df_1, na.rm = TRUE) # mean of test and retest for each person.
df_1$sd_within_i <- abs(df_1$test - df_1$retest) / sqrt(2)
df_1$cv_i <- (df_1$sd_within_i / df_1$mean_i) * 100

# Test-Retest CV
mean(df_1$cv_i, na.rm = TRUE)
sd(df_1$cv_i, na.rm = TRUE)
# 95% CI:
mean(df_1$cv_i, na.rm = TRUE) + c(-1,1)*1.96*sd(df_1$cv_i, na.rm = TRUE)/sqrt(17)
# 2.195750 5.014306 # check


```


## Exercises

Solutions to the exercises can be found 
[here](https://github.com/jdegenfellner/Script_QM3_ZHAW/tree/main/Solutions_Exercises).

### [E] Exercise 1 {#exercise1}

Use the $50$ ROM measurements from Peter and Mary (see above). 

- Introduce biases of $5$, $15$ and $35$ degrees in the data set. Mary should have systematically 
  higher values than Peter.
- Show in R that the correlation does not change.

### [M] Exercise 2 {#exercise2}

Consider the $50$ ROM measurements from Peter and Mary (see above) on the non affected side (nas).

- Regress Mary's measurements on Peter's measurements using the simple
  linear regression model in R.
- How precisely can we predict Mary's measurements from Peter's, i.e. analyse the resdiuals.
- Calculate the mean absolute deviation MAD, i.e. 
  $$MAD = \frac{1}{n} \sum_{i=1}^n |ROM_i - \widehat{ROM}_i|$$
  where $ROM_i$ is the true value and $\widehat{ROM}_i$ is the model-predicted value.

### [E] Exercise 3 {#exercise3}

- Draw the model hierarchy for the first model [above](#ICC) (ICC1).

### [M] Exercise 4 {#exercise4}

Go back to the [section](#badnews) about the ICC and the example of the HADS-A.

- Find your own score, maybe one that is used in your field.
- Find the miminally clinically important difference (MCID) for this score.
- Execute the code above with your score and the MCID and see if 
  the results are similar to the HADS-A example.

### [M] Exercise 5 {#exercise5}

Consider the Bayesian model for the ICC2 and ICC3 above.

- Draw the model hierarchy for this model.

### [H] Exercise 6 {#exercise6}

Read the following sections in the book *Measurement in Medicine*:

- 5.6.2.2 Bland-Altman method
- 8.5.3 Smallest detectable change (a change beyond the measurement error)
- 8.5.4.1 The concept of minimal important change
- Be forgiving with the dichotomous writing style.

### [M] Exercise 7 {#exercise7}

Go back to the 5 degree biased $ROMs$ from Peter and Mary.

- Fit the Frequentist models for ICC1 and ICC2/3 again using `lmer`.
- Perform model diagnostics (`check_model`) for both models.
- Also compare residuals and posterior predictive checks (PPC). You can
  plot the diagnostics separately, for example: `check_model(mod1, check = "pp_check")`.
- What do you think about the model fit for the two models?
- Which model do you prefer?

### [H] Exercise 8 {#exercise8}

Introduce some outliers in the $ROMs$ of Peter and Mary (50 participants).

- How do the outliers influence the ICC?
- How do the outliers influence the model fit of the underlying statistical model?

### [M] Exercise 9 {#exercise9}

Consider the example reliability study [above](#example_RELStudy).

- According to this [paper](https://www.sciencedirect.com/science/article/pii/S1556370716000158?via%3Dihub), 
  did the authors choose the right ICC (ICC1k)?
  What can the chosen ICC be used for?
- Verify the ICC for the "Raw elongation (mm)" in Table 1 of the paper.

### [E] Exercise 10 {#exercise10}

We want to see where formula 5.7 for the SEM in the book comes from.
 $$SEM = SD_{pooled} \cdot \sqrt{1 - ICC}$$

- We start with the formula for the (overall) ICC:
$$ICC = \frac{\sigma_{\eta}^2}{\sigma_{\eta}^2 + \sigma_{\varepsilon}^2}$$

- $SEM = \sqrt{\sigma_{\varepsilon}^2}$

- Show that the formula for the SEM is equivalent to the one in 
  the book by rewriting the ICC formula.

## eLearning

In this eLearning section, please study (the whole) chapter 6 on 
[validity](https://en.wikipedia.org/wiki/Validity_(statistics)) in the book (pages 150-196).
Here are some thoughts on the chapter:

- p. 153: "Existing knowledge about the construct should drive the hypotheses."
  This is an important point. It seems a good idea to apply our Baysian knowledge
  here. Often simply correlation tests or $t$-tests are used in this area. 
  When doing so, it would be advantageous to define priors and use the Bayesian
  analysis we have covered previously.
- p. 154: "They emphasize that the crucial ingredient of validity involves the
  causal role of the construct in determining what value the measurement 
  outcomes will take..." In my opinion, all these analyses should be done using 
  causal inference. Richard McEleath's book contains a lot of information on this.
  But granted: We are at the very beginning on this topic, in the lecture and in wide
  areas of health-research.
- You should be familiar with the different definitions of validity: 
  face, content, concurrent, predictive, structural, convergent, discriminant, 
  discriminiative validity.
- p. 181 at the top: As you can imagine, this is a very important piece of information.
  We are not here to play "significance"-bingo with "hypotheses". Ideally, 
  we are searching for the truth and want to answer real questions 
  (which is unfortunately not always attainable).
- Figure 6.8. on p. 195. is worth remembering.

Additionally, look for at least two reliability and validity articles in your field
(either Reliabiilty alone or Reliability and Validity together) and try to understand what
they did. Did the authors provide code and data to aid reproducibility?

